Adversarial Variational Optimization of\\ Non-Differentiable Simulators

### 2018-01-31 icml/avo.tex
save

\begin{figure}[h]
\centering
\includegraphics[width=0.48\textwidth]{figures/poisson.pdf}
\caption{Discrete Poisson model with unknown mean.
 ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after training. For both $\gamma=0$ and $\gamma=1$, the distributions correctly concentrate their density around
            the true value $\log(\lambda^*)$. Entropic regularization ($\gamma=1$) results in a tighter density.
 ({\it Top right}) Model distributions $\qxpsi$ after training. This plot shows that the resulting parametrizations of the simulator closely reproduce the true distribution.
 ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
 }\label{fig:poisson}
\end{figure}

@@

The observed data is sampled from a Poisson with mean $\lambda^* = 7$.
Algorithm~\ref{alg:avo} is run for 500 epochs with mini-batches of size $M=64$
and the following configuration. For the critic $d$, we use a 3-layer MLP with 10
hidden nodes per layer and ReLU activations. At each epoch, Adam is run for
$n_\text{critic}=5$ iterations with a step size $\alpha=0.01$ and otherwise default parameters.
Its inner first and second moment vectors are reset at
each outer epoch in order to avoid building momentum in staled directions.  For
estimating $\lambda^*$, we parametrize $\bftheta$ as $\log(\lambda)$ and use a univariate Gaussian proposal distribution
$q(\bftheta|\bfpsi)$ initialized with a mean at $\log(5)$ and unit variance. At
each epoch, parameters $\bfpsi$ are updated by taking one Adam step, with step size
$\alpha=0.01$. The gradient penalty coefficient is set to
$\lambda=0.025$, and the entropy penalty is evaluated at both $\gamma=0$ and $\gamma=1$.

The top left plot in Figure~\ref{fig:poisson} illustrates the resulting proposal
distributions $q(\bftheta|\bfpsi)$ after AVO.  For
both $\gamma=0$ and $\gamma=1$, the proposal distributions correctly concentrate
their density around the true parameter value $\log(\lambda^*) = 1.94$. Under
the effect of entropic regularization,
the proposal distribution for $\gamma=1$ concentrates its mass tightly,
yielding in this case precise inference.  The top right plot compares the
model distributions to the true distribution.  As theoretically expected from
adversarial training, we see that the resulting distributions align with
the true distribution, with in this case visually slightly better results for the penalized
model.  The bottom plot of Figure~\ref{fig:poisson} shows empirical estimates
of $-U_d$ with respect to the epoch number. For both $\gamma=0$ and $\gamma=1$,
the curves quickly fall towards $0$, which indicates that
$\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)}
[d(\tilde{\mathbf{x}};\bfphi)] \approx \mathbb{E}_{\mathbf{x} \sim
p_r(\mathbf{x})} [d(\mathbf{x};\bfphi)]$ and that the critic cannot distinguish
between true and model data. This confirms that adversarial variational optimization
works despite the discreteness of the data and lack of access to the
density $p(\bfx | \bftheta)$ or its gradient.

@@

As a second toy example, we consider a generator producing
5-dimensional continuous data, as originally specified in Section 4.2 of
\cite{cranmer2015approximating}.
% More specifically, we consider the following
% generative process:
% \begin{itemize}
% \item $\bfz = (z_0, z_1, z_2, z_3, z_4)$, such that \\
% $z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
% $z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
% $z_2 \sim {\text{Mixture}}[\frac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \\  \frac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5)]$,
% $z_3 \sim {\text{Exponential}(\lambda=3)}$, and
% $z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
% \item $\mathbf{x} = R  \bfz$, where $R$ is a fixed
% semi-positive definite $5 \times 5$ matrix defining a fixed projection
% of $\bfz$ into the observed space.
% \end{itemize}
Again, AVO does not have access to the density or its gradient, only samples from the generative model.
We consider observed data generated at the nominal values $\bftheta^* = (\alpha^*=1,\beta^*=-1)$.
The simulator parameters are modeled with a factored  (mean field) Gaussian
proposal distribution $q(\bftheta|\bfpsi) = q(\alpha|\bfpsi) q(\beta|\bfpsi)$, where each component was
initialized with zero mean and unit variance.
Hyper-parameters are set to $M=64$, $n_\text{critic}=100$, $\lambda=0.025$, $\gamma=10$ and
Adam configured with a step size $\alpha=0.01$ and otherwise default parameters.
The architecture of the critic is the same as in the previous example.

Starting with a proposal distribution $q(\bftheta|\bfpsi)$ largely spread over
the parameter space, as illustrated in the left plot of Figure~\ref{fig:multi},
AVO quickly converges towards a proposal distribution whose density
concentrates around the nominal values $\bftheta^*$, as shown in the right plot of Figure~\ref{fig:multi}.
Overall, this example further illustrates and confirms the ability of adversarial
variational optimization for inference with multiple parameters and multidimensional
data, where reliable approximations of $p(\mathbf{x}|\bftheta)$ in a traditional
MLE setting would otherwise be difficult to construct.

@@

More generally, likelihood-free
inference is intimately tied to a class of algorithms that can be framed as
density estimation-by-comparison, as reviewed in \cite{2016arXiv161003483M}. In most cases, these
inference algorithms are formulated as an iterative two-step process where the
model distribution is first compared to the true data distribution and then
updated to make it more comparable to the latter. Relevant work in this
direction includes those that rely on a classifier to estimate the discrepancy between the observed data and the model distributions
\cite{gutmann2012noise,cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D,gutmann2017likelihood,rosca2017variational}.
% Also of direct relevance in the likelihood-free setup, Hamiltonian ABC~\cite{meeds2015hamiltonian}
% estimate gradients with respect
% to $\bftheta$ through finite differences from multiple forward passes of the
% simulator with variance reduction strategies based on controlling the source of
% randomness used for the latent variable $\bfz$.

@@

This work has also many connections to work on variational inference, in which
the goal is to optimize the recognition model $q(\bfz, \bftheta|\bfpsi)$ so that
it is close to the true posterior $p(\bfz, \bftheta |\mathbf{x})$. There have
been efforts to extend variational inference to intractable likelihoods;
however, many require restrictive assumptions.  In \cite{tran2017variational},
the authors consider Variational Bayes with an Intractable Likelihood (VBIL). In
that approach ``the only requirement is that the intractable likelihood can be
estimated unbiasedly.'' In the case of simulators, they propose to use the
ABC-likelihood with an $\epsilon$-kernel. The ABC likelihood is only unbiased as
$\epsilon \to 0$,   thus this method inherits the drawbacks of the
ABC-likelihood including the choice of summary statistics and the inefficiency
in evaluating the ABC likelihood for high-dimensional data and small $\epsilon$.
More recently, \cite{2017arXiv170208896T} adapted variational inference to
hierarchical implicit models defined on simulators. In this work, the authors
step around the intractable likelihoods by reformulating the optimization of the
ELBO in terms of a neural and differentiable approximation $r$ of the log-likelihood
ratio $\log \tfrac{p}{q}$.

@@

The top left plot in Figure~\ref{fig:weinberg} illustrates the resulting
proposal distributions $q(\bftheta|\bfpsi)$ for $\gamma=0$ and $\gamma=5$ after
AVO. We see that the distributions both arrive
in the neighborhood of $\bftheta^*$, with a density  more highly concentrated for
$\gamma=5$ than for $\gamma=0$.  Despite these differences and the relative
distance with $\bftheta^*$, both models closely match the observed data, as shown
in the top right plot of  Figure~\ref{fig:weinberg}, with again slightly better
results for the entropy penalized model. This suggests either a relatively flat
landscape around $\bftheta^*$ or that the observed data can in this case also be
reproduced with the  predictive distribution $\qxpsi$.
Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that for both
$\gamma=0$ and $\gamma=5$ the variational upper bound $-U_d$ quickly fall
towards $0$, which indicates  convergence towards a distribution that the critic
cannot distinguish from the true distribution.

>>>>

\begin{figure}[h]
\centering
\includegraphics[width=0.48\textwidth]{figures/poisson.pdf}
\caption{Discrete Poisson model with unknown mean.
 ({\it Top left}) Proposal distributions $q(\bftheta|\bfpsi)$ after training. For both $\gamma=0$ and $\gamma=5$, the distributions correctly concentrate their density around
            the true value $\log(\lambda^*)$. Entropic regularization ($\gamma=5$) results in a tighter density.
 ({\it Top right}) Model distributions $\qxpsi$ after training. This plot shows that the resulting parametrizations of the simulator closely reproduce the true distribution.
 ({\it Bottom}) Empirical estimates of the variational upper bound $U_d$ as optimization progresses.
 }\label{fig:poisson}
\end{figure}

@@

The observed data is sampled from a Poisson with mean $\lambda^* = 7$.
Algorithm~\ref{alg:avo} is run for 500 epochs with mini-batches of size $M=64$
and the following configuration. For the critic $d$, we use a 3-layer MLP with 10
hidden nodes per layer and ReLU activations. At each epoch, Adam is run for
$n_\text{critic}=10$ iterations with a step size $\alpha=0.005$ and $\beta_1=\beta_2=0.05$.
% Its inner first and second moment vectors are reset at
% each outer epoch in order to avoid building momentum in staled directions.
For
estimating $\lambda^*$, we parametrize $\bftheta$ as $\log(\lambda)$ and use a univariate Gaussian proposal distribution
$q(\bftheta|\bfpsi)$ initialized with a mean at $\log(10)$ and unit variance. At
each epoch, parameters $\bfpsi$ are updated by taking one Adam step and the same hyper-parameters as for the critic.
The gradient penalty coefficient is set to
$\lambda=0.1$, and the entropy penalty is evaluated at both $\gamma=0$ and $\gamma=5$.

The top left plot in Figure~\ref{fig:poisson} illustrates the resulting proposal
distributions $q(\bftheta|\bfpsi)$ after AVO.  For
both $\gamma=0$ and $\gamma=5$, the proposal distributions correctly concentrate
their density around the true parameter value $\log(\lambda^*) = 1.94$. Under
the effect of entropic regularization,
the proposal distribution for $\gamma=5$ concentrates its mass tightly,
yielding in this case precise inference.  The top right plot compares the
model distributions to the true distribution.  As theoretically expected from
adversarial training, we see that the resulting distributions align with
the true distribution, with in this case visually slightly better results for the penalized
model.  The bottom plot of Figure~\ref{fig:poisson} shows empirical estimates
of $-U_d$ with respect to the epoch number. For both $\gamma=0$ and $\gamma=5$,
the curves fall towards $0$, which indicates that
$\mathbb{E}_{\tilde{\mathbf{x}} \sim p(\mathbf{x}|\bftheta)}
[d(\tilde{\mathbf{x}};\bfphi)] \approx \mathbb{E}_{\mathbf{x} \sim
p_r(\mathbf{x})} [d(\mathbf{x};\bfphi)]$ and that the critic cannot distinguish
between true and model data. This confirms that adversarial variational optimization
works despite the discreteness of the data and lack of access to the
density $p(\bfx | \bftheta)$ or its gradient.

@@

As a second toy example, we consider a generator producing
5-dimensional continuous data, as originally specified in Section 4.2 of
\cite{cranmer2015approximating}.
% More specifically, we consider the following
% generative process:
% \begin{itemize}
% \item $\bfz = (z_0, z_1, z_2, z_3, z_4)$, such that \\
% $z_0 \sim {\cal N}(\mu=\alpha, \sigma=1)$,
% $z_1 \sim {\cal N}(\mu=\beta, \sigma=3)$,
% $z_2 \sim {\text{Mixture}}[\frac{1}{2}\,{\cal N}(\mu=-2, \sigma=1), \\  \frac{1}{2}\,{\cal N}(\mu=2, \sigma=0.5)]$,
% $z_3 \sim {\text{Exponential}(\lambda=3)}$, and
% $z_4 \sim {\text{Exponential}(\lambda=0.5)}$;
% \item $\mathbf{x} = R  \bfz$, where $R$ is a fixed
% semi-positive definite $5 \times 5$ matrix defining a fixed projection
% of $\bfz$ into the observed space.
% \end{itemize}
Again, AVO does not have access to the density or its gradient, only samples from the generative model.
We consider observed data generated at the nominal values $\bftheta^* = (\alpha^*=1,\beta^*=-1)$.
The simulator parameters are modeled with a factored  (mean field) Gaussian
proposal distribution $q(\bftheta|\bfpsi) = q(\alpha|\bfpsi) q(\beta|\bfpsi)$, where each component was
initialized with zero mean and unit variance.
Hyper-parameters are set to $M=64$, $n_\text{critic}=10$, $\lambda=0.1$, $\gamma=1$ and
Adam configured with a step size $\alpha=0.005$ and $\beta_1=\beta_2=0.05$.
The architecture of the critic is the same as in the previous example.

Starting with a proposal distribution $q(\bftheta|\bfpsi)$ largely spread over
the parameter space, as illustrated in the left plot of Figure~\ref{fig:multi},
AVO converges towards a proposal distribution whose density
concentrates around the nominal values $\bftheta^*$, as shown in the right plot of Figure~\ref{fig:multi}.
Overall, this example further illustrates and confirms the ability of adversarial
variational optimization for inference with multiple parameters and multidimensional
data, where reliable approximations of $p(\mathbf{x}|\bftheta)$ in a traditional
MLE setting would otherwise be difficult to construct.

@@

More generally, likelihood-free
inference is intimately tied to a class of algorithms that can be framed as
density estimation-by-comparison, as reviewed in \cite{2016arXiv161003483M}. In most cases, these
inference algorithms are formulated as an iterative two-step process where the
model distribution is first compared to the true data distribution and then
updated to make it more comparable to the latter. Relevant work in this
direction includes those that rely on a classifier to estimate the discrepancy between the observed data and the model distributions
\cite{gutmann2012noise,cranmer2015approximating,cranmer2016experiments,2016arXiv161110242D,gutmann2017likelihood,rosca2017variational}.
Also of direct relevance in the likelihood-free setup, Hamiltonian ABC~\cite{meeds2015hamiltonian}
estimate gradients with respect
to $\bftheta$ through finite differences from multiple forward passes of the
simulator with variance reduction strategies based on controlling the source of
randomness used for the latent variable $\bfz$.

@@

This work has also many connections to work on variational inference, in which
the goal is to optimize the recognition model $q(\bfz, \bftheta|\bfpsi)$ so that
it is close to the true posterior $p(\bfz, \bftheta |\mathbf{x})$. There have
been efforts to extend variational inference to intractable likelihoods;
however, many require restrictive assumptions.  In \cite{tran2017variational},
the authors consider Variational Bayes with an Intractable Likelihood (VBIL). In
that approach ``the only requirement is that the intractable likelihood can be
estimated unbiasedly.'' In the case of simulators, they propose to use the
ABC-likelihood with an $\epsilon$-kernel. The ABC likelihood is only unbiased as
$\epsilon \to 0$,   thus this method inherits the drawbacks of the
ABC-likelihood including the choice of summary statistics and the inefficiency
in evaluating the ABC likelihood for high-dimensional data and small $\epsilon$.
More recently, \cite{2017arXiv170208896T} adapted variational inference to
hierarchical implicit models defined on simulators. In this work, the authors
step around the intractable likelihoods by reformulating the optimization of the
ELBO in terms of a neural and differentiable approximation $r$ of the log-likelihood
ratio $\log \tfrac{p}{q}$, thereby effectively using the same core principle as used
in GANs~\cite{2016arXiv161003483M}.

@@

The top left plot in Figure~\ref{fig:weinberg} illustrates the resulting
proposal distributions $q(\bftheta|\bfpsi)$ for $\gamma=0$ and $\gamma=5$ after
AVO. We see that the distributions both arrive
in the neighborhood of $\bftheta^*$, with a density  more highly concentrated for
$\gamma=5$ than for $\gamma=0$.  Despite these differences and the relative
distance with $\bftheta^*$, both models closely match the observed data, as shown
in the top right plot of  Figure~\ref{fig:weinberg}, with again slightly better
results when using entropic regularization. This suggests either a relatively flat
landscape around $\bftheta^*$ or that the observed data can in this case also be
reproduced with the  predictive distribution $\qxpsi$.
Finally, the bottom plot of Figure~\ref{fig:weinberg} shows that for both
$\gamma=0$ and $\gamma=5$ the variational upper bound $-U_d$ quickly fall
towards $0$, which indicates  convergence towards a distribution that the critic
cannot distinguish from the true distribution.
